### 2008-09-05 trunk/Papers/rmrs-08/rmrs.tex
language more concise and partial conclusion added

\rmrs\ as presented here combines similarities to earlier definitions
of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance constraints
\cite{egg:etal:2001}.  Compared to earlier versions of \rmrs, we write
$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
$\ARG_\N$ instead of $\ARG_n$.  Furthermore, we can express the
elementary predication $\ep{l}{a}{P(v)}$ with the atoms
$\{\ep{a}{l}{P}, \ARG_{\{0\}}(l,v)\}$; note that the positions of the
anchor and label have changed.  In other words, the syntax defined
here differs in some details, but is more general. The primary
differences to dominance constraints are the distinction of labels and
anchors to support EP conjunctions directly, the separation of
atoms for specifying the labelling and the parent-child relations, and
the use of $\qeq$ instead of dominance.  One could thus see our
version of \rmrs\ as a redefinition of classical \rmrs\ in the
dominance constraints style, which will simplify the formal work
below.

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$.  The models of the \rmrs\ will then be such trees, and we
can define notions such as satisfaction and entailment over these
models in the usual way.  Satisfaction corresponds to the relation
between an \rmrs\ and a fully specific semantic representation that it
describes, whereas entailment corresponds to the relation between
a more and a less specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic; in this case, $\tau$ is simply the structure tree of
the predicate logic formula.  Other base languages may require
different mappings between formulas and trees.  We assume that
$\Sigma$ is a ranked signature consisting of the symbols of the base
language with appropriate arities.  To encode predicate logic, we
assume the unary constructor $\neg$ and binary constructors $\wedge$,
$\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
x$ for all base-language variables $x$; all the constants and
base-language variables as nullary constructors; and, for each
predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
For other base languages, a different signature could be chosen; we
only assume for convenience that the signature contains a binary
constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
for the set of all constructors in $\Sigma$ that have arity $i$.

@@

The {\em satisfaction} relation between an {\sc rmrs} $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, we assume a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
maps the base variables in the \rmrs\ to nullary constructors
representing variables in the base language.  This function enables us
to map two variables that seem different in the \rmrs\ to the same
base-language variable, introducing e.g.\ semantic dependencies that a
POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
in (\ref{eg:cat-pos}b) to $y$ to specify the restrictor of the
quantifier.  Finally, we assume a relation $\sigma \subseteq \pred
\times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
a set of constructors from $\Sigma$.  This relation allows us to
underspecify lexical ambiguities; for instance,
$\sigma(\mathrm{\_cat\_n})$ might contain the symbol $\mathrm{\_cat\_n\_1}$
in (\ref{eg:cat}c) along with symbols for other word senses.

@@

Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
This \rmrs\ is normal, and it is of a particularly simple form because
it consists only of elementary predications.  The \rmrs\ is satisfied
(among others) by the tree in Fig.~\ref{fig:1} together with some
particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
map the variables $l_1$ and $a_1$ to the root node; $l_{41}$ and
$l_{42}$ to its second child (labeled with ``$\wedge$''); $a_{41}$ to
the first child of \emph{that} node (i.e.\ the node 21, labelled with
``fat'') and $a_{42}$ to the node 22; and so forth.  $g$ will map $x$
and $x_3$ to $x$, $e'$ to $e$, and so on.  And $\sigma$ will map each
\rmrs\ predicate symbol (which represents a word) to the set of its
word senses, e.g.\ \_cat\_n to a set containing \_cat\_n\_1 and
possibly others.  It is then easy to verify that every single atom in
the \rmrs\ is satisfied -- most interestingly, the EPs
$\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``the fat black cat chased the
dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
the restrictor of the universal quantifier.  It is still a model of
(\ref{eg:cat-pos}b) -- all the EPs are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula) in the space
between the labels $l_{41}$ and $l_{42}$ and the anchors $a_{41}$ and
$a_{42}$.  The \rmrs\ didn't specify that this material should be
there; but it also didn't specify that it can't be there, and indeed
there is no way in which the \rmrs\ can enforce that no extra material
can be added.

@@

In the context of robust semantic processing, this is a desirable
feature of \rmrs, because it means that when we enrich an \rmrs\
obtained from a shallow processor with more semantic information, we
don't change the set of models; we only \emph{restrict} the set of
models further and further towards the semantic representation we are
trying to reconstruct.  However, the fact that all satisfiable \rmrs s
have an infinite number of models is a problem for algorithms that try
to decide satisfiability or compute the fully specific semantic
representations -- how can this be done in finite time?

In this paper, we follow the dominance constraint tradition in
defining a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
s in solved form are guaranteed to be satisfiable, and thus each
solved form represents an infinite class of models.  However, we will
set it up in such a way that each satisfiable \rmrs\ has only a finite
number of solved forms which partition the space of possible models
into classes with only ``irrelevant'' differences as above.  A solver
can then enumerate the solved forms rather than all models.

Intuitively, we expect that an \rmrs\ in solved form is fully
specified with respect to scope ambiguities and the predicate-argument
structure, all variable equalities and inequalities have been
resolved, and only contains lexical ambiguities that are ``harmless'',
inthat any choice of a constructor for the predicate symbol will
satisfy the \rmrs.  This is made precise below.

@@

In the earlier \rmrs\ literature, each term of an \rmrs\ is equipped
with a \emph{sort}.  For instance, individual variables are typically
assigned the sort $x$, event variables the sort $e$, and holes the
4sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
for events in the past tense), and all the sorts arranged in a sort
hierarchy ${\cal S}$.

While we have defined \rmrs\ without sorts above to simplify the
presentation, it is straightforward to add sorts to the definition.
For this, one assumes that the signature $\Sigma$ is sorted, i.e.\
assigns a sort $s_1\times\ldots s_n\rightarrow s$ to each constructor,
where $n$ is the constructor's arity (possibly zero) and
$s,s_1,\ldots,s_n\in {\cal S}$ are atomic sorts.  We can then
restrict the models of \rmrs\ to trees that are \emph{well-sorted} in
the usual sense, i.e.\ those in which we can infer a sort for each
subtree, and require that the variable assignment functions likewise
respect the sorts.  If we then modify Def.~\ref{defn:solved-forms}
such that the constructor $p$ of sufficiently high arity is also
consistent with the sorts of the known arguments -- i.e., if $p$ has
sort $s_1 \times \ldots \times s_n \rightarrow s$ and the \rmrs\
contains an atom $\ARG_{\{k\}}(Y,i)$ and $i$ is of sort $s'$, then
$s'$ is a subsort of $s_k$ --, all the above propositions about solved
forms remain true.

Another extension that is worth discussing is the idea of replacing
the qeq atom, $X \qeqdom Y$, by ordinary dominance $X \dom Y$ in the
sense of dominance constraints, which is satisfied iff there is a path
from $\alpha(X)$ to $\alpha(Y)$ in the tree.  Dominance is clearly a
less restrictive condition than qeq, but it has been shown that the
distinction between the two relations is irrelevant in \mrs s that were
derived from a corpus by a deep grammar \cite{FucKolNieTha04}.  In our
above definition of \rmrs, it is possible to replace qeq by dominance;
this changes \rmrs\ to a different logic (let's call it
Dominance-\rmrs), for which however all the above propositions about
solved forms still hold if ``qeq'' is systematically replaced by
``dominance''.

>>>>

\rmrs\ as presented here combines similarities to earlier definitions
of \rmrs\ (e.g., \cite{copestake:2003}) as illustrated in
(\ref{eg:cat}c) and
to dominance constraints
\cite{egg:etal:2001}.  We
$\ARG_{\{i\}}$ (with a singleton set $S$) instead of $\ARG_i$, and
$\ARG_\N$ instead of $\ARG_n$.  Furthermore, the
elementary predication $\ep{l}{a}{P(v)}$ is a notational variant of
$\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.
In other words, the syntax defined
here differs in some details, but is more general. The primary
differences to dominance constraints are the distinction of labels and
anchors that will separate the
labelling and the parent-child relations, and
the use of $\qeq$ instead of dominance.  Redefining
\rmrs\ in the
style of dominance constraints will simplify the formal work
below.

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$ (see Figure~\ref{fig:1}).  So models of the \rmrs\ are
such trees, and we 
can define satisfaction and entailment over these
models in the usual way.  Satisfaction corresponds to 
an \rmrs\ being a partial description of the (unique) fully specific
semantic representation that the model corresponds to, whereas
entailment corresponds to the relation between
a more and a less specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic; in this case, $\tau$ is simply the structure tree of
the predicate logic formula.  We assume that
$\Sigma$ is a ranked signature consisting of predicate logic's
symbols:
a unary constructor $\neg$ and binary constructors $\wedge$,
$\rightarrow$, etc.; the binary constructors $\forall x$ and $\exists
x$ for all base-language variables $x$; all the constants and
base-language variables as nullary constructors; and, for each
predicate symbol $P$ of arity $n$, a constructor $P$ of arity $n$.
Other base languages may require a different signature $\Sigma$ and/or
a different mapping between formulas and trees.  We
only assume for convenience that the signature contains a binary
constructor $\wedge$ to represent conjunction.  We write $\Sigma_i$
for the set of all constructors in $\Sigma$ that have arity $i$.

@@

The {\em satisfaction} relation between an {\sc rmrs} $\varphi$ and a
finite constructor tree $\tau$ is defined in terms of several
assignment functions.  First, we assume a node variable assignment
function $\alpha: \nvar \rightarrow D(\tau)$ that maps the node
variables in an \rmrs\ to the nodes of $\tau$.  Second, we assume a
base language assignment function $g:\ovar \rightarrow \Sigma_0$ that
maps the base variables in the \rmrs\ to nullary constructors
representing variables in the base language.  This function enables us
to map two variables that seem different in the \rmrs\ to the same
base-language variable, introducing e.g.\ semantic dependencies that a
POS tagger didn't catch.  For instance, $g$ might map $x_4$ and $x_5$
in (\ref{eg:cat-pos}b) to $y$ to specify a variable binding.  Finally,
we assume a relation $\sigma \subseteq \pred 
\times \Sigma_{\geq 1}$ that maps each {\sc rmrs} predicate symbol to
a set of constructors from $\Sigma$.  This relation allows us to
underspecify lexical ambiguities; for instance,
$\sigma(\mathrm{\_cat\_n})$ might contain the symbol $\mathrm{\_cat\_n\_1}$
in (\ref{eg:cat}c) along with symbols for other word senses.

@@

Let's look at the \rmrs\ in (\ref{eg:cat-pos}b) by way of example.
This \rmrs\ is satisfied
by the tree in Fig.~\ref{fig:1} (among others) together with some
particular $\alpha$, $g$, and $\sigma$.  The assignment $\alpha$ will
map the variables $l_1$ and $a_1$ to the root node; $l_{41}$ and
$l_{42}$ to its second child (labeled with ``$\wedge$''); $a_{41}$ to
the first child of \emph{that} node (i.e.\ the node 21, labelled with
``fat'') and $a_{42}$ to the node 22; and so forth.  $g$ will map $x$
and $x_3$ to $x$, $e'$ to $e$, and so on.  And $\sigma$ will map each
\rmrs\ predicate symbol (which represents a word) to the set of its
word senses, e.g.\ \_cat\_n to a set containing \_cat\_n\_1 and
possibly others.  It is then easy to verify that every single atom in
the \rmrs\ is satisfied -- most interestingly, the EPs
$\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``every fat black cat chased some
dog'', differs from the model in Fig.~\ref{fig:1} in the subtree below
the restrictor of the universal quantifier.  It is still a model of
(\ref{eg:cat-pos}b)---all the EPs are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula).
The \rmrs\ didn't specify that this material should be
there; but it also didn't specify that it can't be there, and indeed
there is no way in which the \rmrs\ can enforce that no extra material
can be added.

@@

In the context of robust semantic processing, this is a desirable
feature of \rmrs, because it means that when we enrich an \rmrs\
obtained from a shallow processor with more semantic information, we
don't change the set of models; we only \emph{restrict} the set of
models further and further towards the semantic representation we are
trying to reconstruct.  However, the fact that all satisfiable \rmrs s
have an infinite number of models is a problem for algorithms that try
to decide satisfiability or compute the fully specific semantic
representations---how can this be done in finite time?

In this paper, we follow the dominance constraint strategy for
resolving this dilemma; we
define a fragment of \rmrs s called \emph{solved forms}.  All \rmrs
s in solved form are guaranteed to be satisfiable, and thus each
solved form represents an infinite class of models.  However, we will
set it up in such a way that each satisfiable \rmrs\ has only a finite
number of solved forms which partition the space of possible models
into classes with only ``irrelevant'' differences as above.  A solver
can then enumerate the solved forms rather than all models.

Intuitively, we expect that an \rmrs\ in solved form is fully
specified with respect to scope ambiguities and the predicate-argument
structure, all variable equalities and inequalities have been
resolved, and only lexical sense ambiguities remain.  This is made
precise below.

@@

In the earlier \rmrs\ literature, each term of an \rmrs\ is equipped
with a \emph{sort}.  For instance, individual variables are typically
assigned the sort $x$, event variables the sort $e$, and holes the
sort $h$; sometimes subsorts of these are considered (e.g.\ $e_{PAST}$
for events in the past tense), and all the sorts arranged in a sort
hierarchy ${\cal S}$.

While we have defined \rmrs\ without sorts to simplify the
presentation, it is straightforward to add sorts to the definition.
For this, one assumes that the signature $\Sigma$ is sorted, i.e.\
assigns a sort $s_1\times\ldots s_n\rightarrow s$ to each constructor,
where $n$ is the constructor's arity (possibly zero) and
$s,s_1,\ldots,s_n\in {\cal S}$ are atomic sorts.  We can then
restrict the models of \rmrs\ to trees that are \emph{well-sorted} in
the usual sense, i.e.\ those in which we can infer a sort for each
subtree, and require that the variable assignment functions likewise
respect the sorts.  If we then modify Def.~\ref{defn:solved-forms}
such that the constructor $p$ of sufficiently high arity is also
consistent with the sorts of the known arguments---i.e., if $p$ has
sort $s_1 \times \ldots \times s_n \rightarrow s$ and the \rmrs\
contains an atom $\ARG_{\{k\}}(Y,i)$ and $i$ is of sort $s'$, then
$s'$ is a subsort of $s_k$---all the above propositions about solved
forms remain true.

Another extension is to replace
the qeq atom $X \qeqdom Y$ with dominance $X \dom Y$, 
which is satisfied iff there is a path
from $\alpha(X)$ to $\alpha(Y)$ in the tree.  Dominance is clearly a
less restrictive condition than qeq, but it has been shown that the
distinction between the two relations is irrelevant in \mrs s that were
derived from a corpus by a deep grammar \cite{FucKolNieTha04}.  In our
above definition of \rmrs, it is possible to replace qeq by dominance;
this changes \rmrs\ to a different logic (let's call it
Dominance-\rmrs), for which however all the above propositions about
solved forms still hold if ``qeq'' is systematically replaced by
``dominance''.
